Experiences with Planning for Natural Language Generation

### 2009-06-03 trunk/Papers/ci-spark/discussion.tex
changes to section 5

There are both positive and negative conclusions about the state of
planning for modern NLG applications that we can draw from these
experiments.

On the one hand, we found that modern planners are very good at
dealing with the \emph{search} problems that are involved in the
NLG-based planning problems we looked at. In the sentence generation
domain, FF's Enforced Hill-Climbing strategy finds plans corresponding
to 25-word sentences in about a second. It is hard to compare this
number to a baseline because there are no shared benchmark problems;
but FF's search outperforms a greedy search algorithm, and the
performance is competitive with other sentence generators, which
similarly employ special-purpose algorithms. Thus research on search
strategies for planning has paid off: In particular, the Enforced
Hill-Climbing heuristic outperforms the best-first strategy to which
FF 2.3 switches for some problem instances. Similarly, SGPLAN's
performance on the GIVE domain is very convincing and fast enough for
the application.

However, each of the off-the-shelf planners we tested spent
substantial amounts of time on preprocessing. This is most apparent in
the sentence generation domain, where the planners spent almost their
entire runtime on grounding the predicates and operators for some
instances. This effect is much weaker in the GIVE domain, which has a
much smaller number of operators; but as we added dummy predicates,
the ratio of time that the planners spent on preprocessing rose in a
similar way. While the GIVE domain can be carefully defined in such a
way that the number of operators is minimized, this is not possible
for a domain in which the operators encode the different communicative
actions that the NLG system can use. For instance, in the sentence
generation domain, the XTAG planning problem for $k=2$ and $n=5$
consists of about 1000 operators for the different lexicon entries for
all the words in the sentence, some of which take four parameters. It
is not unrealistic to assume a knowledge base with a few hundred
individuals. All this adds up to trillions of ground instances, i.e.\
it is completely infeasible to compute ground instances naively.

This means that while current planners deal well with the search
problems we have looked at, they are still unusable for practical NLG
applications because of the time they spend on preprocessing. In other
words, generation-as-planning research is still not in a much better
position than it was in the 1980s. While we are aware that the time
that a planner invests in preprocessing can pay off during search, we
still suggest that the inability of current planners to scale to
larger domains still limits their usefulness for applications beyond
NLG as well. Furthermore, we feel that preprocessing gets less
research attention than it deserves: If it is scientifically trivial,
we challenge the planning community to come up with implementations
that only ground operators by need; otherwise, we look forward to
publications on this topic. To support this, we offer our planning
domains as benchmarks for future competitions. They are available at
\todo{upload them somewhere}

Finally, we found it very convenient that the recent IPC planning
competitions provide an entry point for selecting and obtaining
current planners. Nevertheless, our experiments exposed several bugs
in the planners we tested, and required us to make changes to the
source code to make them scale to our inputs. We also found that
different planners differ in the variants of PDDL that they can
process. These differences range from fragments of ADL that can be
parsed to sensitivity to the order of declarations and the use of
``objects'' rather than ``individuals'' as the keyword for declaring
the universe. We propose that the case for planning as a mature
technology with professional-quality implementations could be made
more strongly if such discrepancies were harmonized.

>>>>

We can draw both positive and negative conclusions from our experiments about
the state of planning for modern NLG applications. On the one hand, we found
that modern planners are very good at dealing with the \emph{search} problems
that arise in the NLG-based planning problems we investigated; the search
time reported by most planners was often a very small fraction of the total
running time of the planner. In the sentence generation domain, FF's Enforced
Hill-Climbing strategy finds plans corresponding to 25-word sentences in
about a second. It is hard to compare this number to a baseline because there
are no shared benchmark problems, but FF's search outperforms a greedy search
algorithm, and the performance is competitive with other sentence generators,
which similarly employ special-purpose algorithms. Thus research on search
strategies for planning has paid off; in particular, the Enforced
Hill-Climbing heuristic outperforms the best-first strategy to which FF 2.3
switches for some problem instances. Similarly, SGPLAN's performance on the
GIVE domain is very convincing and fast enough for this application.

On the other hand, each of the off-the-shelf planners we tested spent
substantial amounts of time on preprocessing. This is most apparent in the
sentence generation domain, where the planners spent almost their entire
runtime on grounding the predicates and operators for some problem instances.
This effect is much weaker in the GIVE domain, which has a much smaller
number of operators and less interactions between the predicates in the
domain. However, our GIVE experiments also illustrate that altering the
structure of a domain, even minimally, can significantly change a planner's
performance on a problem. For instance, in some of our GIVE experiments with
extra grid positions, increasing the number of buttons in the world, while
keeping the dimensions of the grid fixed, resulted in a significantly larger
search time while the preprocessing time remained essentially unchanged.

While the GIVE domain can be defined in such a way that the number of
operators is minimized, this is not possible for an encoding of a domain in
which the operators model the different communicative actions that the NLG
system can use. For instance, in the sentence generation domain, the XTAG
planning problem for $k=2$ and $n=5$ consists of about 1000 operators for the
different lexicon entries for all the words in the sentence, some of which
take four parameters. It is not unrealistic to assume a knowledge base with a
few hundred individuals. All this adds up to trillions of ground
instances: a set which is completely infeasible to compute naively.

Of course, it would be premature to judge the usefulness of current planners
as a whole, based on just two NLG domains. Nevertheless, we believe that the
structure of our planning problems, which are dominated by large numbers of
operators and individuals, is typical of NLG-related planning problems as a
whole. This strongly suggests that while current planners are able to manage
many of the search problems in the domains we looked at, they are still
unusable for practical NLG applications because of the time they spend on
preprocessing. In other words, the state of generation-as-planning research
is still not in a much better position than it was in the 1980s.

We are also aware that the time a planner invests in preprocessing can pay
off during search, and that such techniques have been invaluable in improving
the overall running time of modern planners. However, we still suggest that
the inability of current planners to scale to larger domains limits their
usefulness for applications beyond NLG as well. Furthermore, we feel that the
problem of preprocessing receives less research attention than it deserves:
if the problem is scientifically trivial then we challenge the planning
community to develop more efficient implementations that only ground
operators by need; otherwise, we look forward to future publications on this
topic. To support this effort, we offer our planning domains as benchmarks
for future research and competitions.\footnote{The PDDL problem generators
for our NLG domains are available at \texttt{http://\todo{TODO}}.}

Finally, we found it very convenient that the recent International Planning
Competitions provide a useful entry point for selecting and obtaining current
planners. Nevertheless, our experiments exposed several bugs in the planners
we tested, which required us to change their source code to make them scale
to our inputs. We also found that different planners that solve the same
class of planning problems (e.g., STRIPS, ADL, etc.) sometimes differ in the
variants of PDDL that they support. These differences range from fragments of
ADL that can be parsed to sensitivity, to the order of declarations and the
use of ``objects'' rather than ``individuals'' as the keyword for declaring
the universe. We propose that the case for planning as a mature technology
with professional-quality implementations could be made more strongly if such
discrepancies were harmonized.

### 2009-06-03 trunk/Papers/ci-spark/domains.tex
changes to section 5

We begin by considering two specific NLG problems: sentence generation in
the sense of \citet{KolSto07}, and the generation of instructions in
virtual environments \citep{ByrKolStrCasDalMooObe09}. For each problem, we
introduce the task and show by example how the task can be viewed as a
planning problem.

@@

However, the problem of deciding whether a given communicative goal can be
achieved with a given grammar is NP-complete \citep{KolStr02}: a naive
search algorithm that computes a derivation top-down takes exponential time
and is clearly infeasible to use in practice. In order to circumvent this
combinatorial explosion, the seminal SPUD system \citep{Stone2003a}, which
first established the idea of integrated TAG-based sentence generation,
used a greedy, but incomplete, search algorithm.
%
To better control the search, \citet{KolSto07} recently proposed an
alternative approach which converts the sentence generation problem into a
planning problem, and solves the transformed search problem using a planner
\citep{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} The resulting planning problem in
this case assumes an initial state containing an atom
$\mathsf{subst}(S,\mathsf{root})$, encoding the fact that a sentence ($S$)
must be generated starting at the node named $\mathsf{root}$ in the TAG
derivation tree, and a second atom $\mathsf{referent}(\mathsf{root},e)$
which encodes the fact that the entire sentence describes the (event)
individual $e$. The elementary trees in the TAG derivation are encoded as
individual planning operators.

@@

Planning plays a central role in the GIVE task. For instance, consider the
example GIVE world shown map in Figure~\ref{fig:give-development-world}. In
this world, the user's task is to pick up a trophy in the top left room.
The trophy is hidden in a safe behind a picture; to access it, the user
must push certain buttons in order to move the picture out of the way, open
the safe, and open doors. The user must navigate the world and perform
these actions in the 3D client; the NLG system must instruct the user on
how to do this. To simplify both the planning and the NLG task, the world
is discretised into a set of tiles of equal size. The user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next tile, provided the path between two tiles is not
blocked. Figure~\ref{fig:give-planning} shows the encoding of some of the
available GIVE domain actions in PDDL syntax. In the example, the shortest
plan to solve the task consists of 108 action steps, with the first few
steps as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{alarm}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{b1, pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

@@

(:action manipulate-alarm-button-off-on
   :parameters (?b - button  ?pos - position  ?alarm - position)
   :precondition 
       (and (state ?b off)  (player-position ?pos)  (position ?b ?pos)
            (controls-alarm ?b ?alarm))
   :effect
       (and (not (state ?b off))  (not (alarmed ?alarm))  (state ?b on)))
\end{verbatim}}%
\end{minipage}
\caption{Simplified PDDL actions for the GIVE domain.}
\label{fig:give-planning}
\end{figure}

>>>>

We begin by considering two specific NLG problems: sentence generation in the
sense of \citet{KolSto07}, and the generation of instructions in virtual
environments \citep{ByrKolStrCasDalMooObe09}. In each case, we introduce the
task and show by example how it can be viewed as a planning problem.

@@

However, the problem of deciding whether a given communicative goal can be
achieved with a given grammar is NP-complete \citep{KolStr02}: a naive
search algorithm that computes a derivation top-down takes exponential time
and is clearly infeasible to use in practice. In order to circumvent this
combinatorial explosion, the seminal SPUD system \citep{Stone2003a}, which
first established the idea of integrated TAG-based sentence generation,
used a greedy, but incomplete, search algorithm.
%
To better control the search, \citet{KolSto07} recently proposed an
alternative approach which converts the sentence generation problem into a
planning problem, and solves the transformed search problem using a planner
\citep{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, which
  implements this conversion.} The resulting planning problem in
this case assumes an initial state containing an atom
$\mathsf{subst}(S,\mathsf{root})$, encoding the fact that a sentence ($S$)
must be generated starting at the node named $\mathsf{root}$ in the TAG
derivation tree, and a second atom $\mathsf{referent}(\mathsf{root},e)$
which encodes the fact that the entire sentence describes the (event)
individual $e$. The elementary trees in the TAG derivation are encoded as
individual planning operators.

@@

Planning plays a central role in the GIVE task. For instance, consider the
example GIVE world shown map in Figure~\ref{fig:give-development-world}. In
this world, the user's task is to pick up a trophy in the top left room.
The trophy is hidden in a safe behind a picture; to access it, the user
must push certain buttons in order to move the picture out of the way, open
the safe, and open doors. The user must navigate the world and perform
these actions in the 3D client; the NLG system must instruct the user on
how to do this. To simplify both the planning and the NLG task, the world
is discretised into a set of tiles of equal size. The user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next tile, provided the path between two tiles is not
blocked. Figure~\ref{fig:give-planning} shows the encoding of some of the
available GIVE domain actions in PDDL syntax. In the example, the shortest
plan to solve the task consists of 108 action steps, with the first few
steps as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{button}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{b1, pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

@@

(:action manipulate-button-off-on
   :parameters (?b - button  ?pos - position  ?alarm - position)
   :precondition 
       (and (state ?b off)  (player-position ?pos)  (position ?b ?pos)
            (controls-alarm ?b ?alarm))
   :effect
       (and (not (state ?b off))  (not (alarmed ?alarm))  (state ?b on)))
\end{verbatim}}%
\end{minipage}
\caption{Simplified PDDL actions for the GIVE domain.}
\label{fig:give-planning}
\end{figure}
